Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-03 paper_jasa.tex
clustering visible plot added
figure 2 caption changed
added references

Statistical graphics research and development is agreeably established now, and has been cemented by several new developments in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{heer:2010} with findings supporting the original results. This body of work provides a framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3.8in]{images/powerplot.pdf} 
   \caption{Probability that the data plot has the smallest probability in a lineup of size $m$. With increasing $p$-value the probability drops -- when it reaches $1/m$ a horizontal line is drawn to emphasize insufficient sensitivity of the test due to the lineup size. 
   %Comparison of probabilities to pick the actual data plot for different size lineups, $m$, and different values for data plot strength, $p_D$. The dashed  line, and grey 95\% pointwise confidence band, shows the probability of picking the data plot, by chance, if $H_0$ is true. }
}
   \label{fig:pval_power}
\end{figure}

@@

By matching the expected values, we therefore get an expression for the $p$-value in the data under $H_1$  as:
\begin{equation}\label{eqn:power_estimate}
%{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
\hh{1 - E[N_m]/K = (m-1)E[p_D] + O(E[p_D^2])}
\end{equation}

Note that the left hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$.

@@

\noindent It was possible to choose more than one reason. The question begged for ``visible trend'' (2) to be the reasons that observers gave for correctly choosing the actual data plot. Either reasons 1 or 3 would be indicative of Type III error. It is interesting to calculate the proportion of observers who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For observers who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??}

>>>>

Statistical graphics research and development is agreeably established now, and has been cemented by several new developments in recent years. Early studies on evaluating how well statistical plots are perceived and read by the human eye \citep{cleveland:1984}, have been repeated and expanded \citep{simkin:1987,spence:1991, heer:2010} with findings supporting the original results. This body of work provides a framework for evaluating new statistical graphics.  In a complementary direction, new research on formalizing statistical graphics with language characteristics makes it easier to abstractly define, compare and contrast data plots. \cite{wilkinson:1999} developed a grammar of graphics that is enhanced by \cite{hadley:2009}. These methods provide a mechanism to abstract the way data is mapped to graphical form. Finally, technology advances make it simple and easy for everyone to draw plots of data, and particularly the existence of software systems, such as R \citep{R}, enable making beautiful data graphics that can be tightly coupled with statistical modeling.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{images/powerplot.pdf} 
   \caption{Probability that the data plot has the smallest probability in a lineup of size $m$. With increasing $p$-value the probability drops -- when it reaches $1/m$ a horizontal line is drawn to emphasize insufficient sensitivity of the test due to the lineup size. 
   %Comparison of probabilities to pick the actual data plot for different size lineups, $m$, and different values for data plot strength, $p_D$. The dashed  line, and grey 95\% pointwise confidence band, shows the probability of picking the data plot, by chance, if $H_0$ is true. }
}
   \label{fig:pval_power}
\end{figure}

@@

By matching the expected values, we therefore get an expression  for the $p$-value $p_D$ in the data 
%under $H_1$  %% no such thing aas $p$-value under H_1 - even though it's just an unfortunate choice of word.
as:
\begin{equation}\label{eqn:power_estimate}
%{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
\hh{1 - E[N_m]/K = (m-1)E[p_D] + O(E[p_D^2])}
\end{equation}

Note that the left hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$. 
\hh{Also note, that this equation yields an estimate of $p_D$ that is bounded by $1/(m-1)$. We should therefore not make the decision of significance depend on this estimate, but only use this equation after the fact. 
Both of these topics are further highlighted in the application section.}

@@

\noindent It was possible to choose more than one reason. The question begged for ``visible trend'' (2) to be the reasons that observers gave for correctly choosing the actual data plot. Either reasons 1 or 3 would be indicative of Type III error. It is interesting to calculate the proportion of observers who choose the correct plot, but use reasons 1 or 3, relative to reason 2. Figure \ref{fig:choice_reason} examines this. For observers who chose the actual data plot, the proportions for the three reasons are shown. It can be seen that the most common reason was ``most different plot'' (1), followed by clustering visible (3), with ``visible trend'' coming in third. \green{What about other??} \blue{Not many people chose other as we see in figure \ref{fig:choice_reason}.}
